Analysing submitted data for OKUM, GAS or MUH-1
========================================================
Thomas Meisel (`r as.character(Sys.Date())`)  
The data for this interlaboratory comparison based certification of property values were analysed by 36 labs following the nested design approached as proposed the IAG certification protocol. Participating labs received 3 packages of OKUM and MUH-1 respectively and one package of GAS. The latter was supplied as a "traceablility" sample and is here used for quality control purposes. It was the task of the labs to prepare two independent sample preparations (i.e. digstions) of each packet and analyse the preparations on two different days. Labs thus should have submitted 12 values (3x2x2 PacketxPrepxDay). 
```{r Data_prep, echo=FALSE, message=FALSE, warning=FALSE}
# Moved this outside the document for easier reading
```
#### importing the data and assigning factors
```{r}
source("Makefile.R")
```

```{r results='hide', include=FALSE}
library(ggplot2) # needed for most plots in this document
#library(devtools) #needed for "order_by"
#install_github("plotflow", "trinker")
library(plotflow) #needed for "order_by" # install_github("plotflow", "trinker")
library(metRology) #for mandel k and h calculations
library(ape) # needed for varcomp (variance component) extraction
library(nlme) # needed for lme
library(plyr) # needed for ddply
library(knitr)
opts_chunk$set(dev="png",dev.args=list(type="cairo"), dpi=96)
options(base64_images = 'inline')



### defining the RM and measurand to be analysed
```{r defining the measurand and RM}
refmat <- 'OKUM' # defining the RM
measurand.name <- 'Ba' # defining the measurand
``` 
```{r defining the variables used for selected RM, include=FALSE}
'%p%' <- function(a, b) {as.character(paste (a, b, sep =""))} # the measureands have different endings depending on which RM: OKUM not extention, MUH extention ".1" and GAS has the extention ".2"
switch(
    refmat,
    GAS = rm1 <- 2,
    MUH = rm1 <- 1,
    OKUM = rm1 <- 0
  )
  if(rm1 > 0) 
  {measurand <- measurand.name %p% '.' %p% rm1
  } else 
  {
    measurand <- measurand.name
  }
```


# measurand selected `r measurand.name` for candiate reference material `r refmat`
```{r general comments}
# Data for certification project was gathered and joined in Excel. The files were exported from Excel as xxxx.csv files to make them universially readable. For this markdown the data is stored in the "root/documents" directory.
# Data is loaded ("GOMGather1.R") and merged ("GOMMerge.R") for GAS, OKUM and MUH-1 are merged together with a methods file ("OKUM.method") into a universal data.frame file named 'GOM'. All of this happens in the "Makefile.R"
```


```{r defining if measurand Major or Trace element}
MorT <- grep(measurand.name, colnames(GOM), fixed=TRUE) # finding the position of the measurand.name in the Columnheaders of dataframe
ifelse(MorT[1]< 21, MorT <- 'M', MorT<-'T') # testing if measurand is a major or trace element/compound (col:5-20 majors)
ifelse(MorT == "T", unit <- 'mg/kg', unit <- 'g/100g') # testing which unit is needed
```

#### defining the plotting theme

```{r mytheme, include=FALSE}
mytheme <- theme_grey() + theme(plot.title = element_text(colour = "black", size = rel(2))) + theme(axis.title.x = element_text(size = rel(1.8)))+ theme(axis.title.y = element_text(size = rel(1.8))) +theme(axis.text.x = element_text(size = rel(1.5))) + theme(axis.text.y = element_text(size = rel(1.5))) + theme(legend.title = element_text(size = rel(1))) +theme(legend.text = element_text(size = rel(0.8)))
```
#### defining the function for plotting methods vs. measurand mass fraction. Sample preparation methods are also marked in the plot.
```{r plot_method function plotting measurand vs. method, include=FALSE, cache=TRUE}
### function plot_method defined here. Automated plot of methods boxplots sorted by increasing method median. Enter with plot_method=('xx') #######
plot_method <- function(x) {
  element <- x
  '%p%' <- function(x, y) {as.character(paste (x, y, sep =""))}
  prep <-  'Prep.'
  method <- 'Method.'
  anal.prep <- prep %p% measurand.name
  anal.method <- method %p% measurand.name
  anal <- GOM[[element]]
  anal.prep <- GOM[[anal.prep]]
  anal.method <- GOM[[anal.method]]
  analyte <- data.frame(GOM$Lab, GOM$names, anal, anal.prep, anal.method )
  analyte <- na.omit(analyte)
  p <- ggplot(reorder_by(anal.method, ~anal, analyte, median), aes(anal.method, anal))+ geom_boxplot(outlier.shape = NA) + geom_jitter(aes(colour=anal.prep), size = 4.5, position = position_jitter(width = .2))
  p + xlab("method") + ylab(unit) + labs(title = measurand.name) + labs(colour = "Prep") + mytheme
}
```


#### defining the function of Youden plots

```{r plot_youd function for youden plot, include=FALSE}
plot_youd <- function(a, y, z) { # a = measurand, y and z ref
  '%p%' <- function(a, b) {as.character(paste (a, b, sep =""))}
  switch(
    y,
    GAS = rm1 <- 2,
    MUH = rm1 <- 1,
    OKUM = rm1 <- 0
  )
  switch(
    z,
    GAS = rm2 <- 2,
    MUH = rm2 <- 1,
    OKUM = rm2 <- 0
  )
  if(rm1 > 0) 
  {RM1 <- a %p% '.' %p% rm1
  } else 
  {
    RM1 <- a
  }
  if(rm2 > 0) 
  {RM2 <- a %p% '.' %p% rm2
  } else 
  {
    RM2 <- a
  }
 if(rm1 > 1)
   {RM1.s <- GOM.sd[[RM1]]} else
     {RM1.s <- GOM.sd[[RM1]]/sd(GOM.median[[RM1]], na.rm = TRUE)}# calculating the normalised standard deviations
 if(rm2 > 1)
   {RM2.s <- GOM.sd[[RM2]]} else
     {RM2.s <- GOM.sd[[RM2]]/sd(GOM.median[[RM2]], na.rm = TRUE)}# calculating the normalised standard deviations
  RM1 <- GOM.median[[RM1]]
  RM1 <- (RM1-median(RM1, na.rm = TRUE))/sd(RM1, na.rm = TRUE) #calculating z-scores
  RM2 <- GOM.median[[RM2]]
  RM2 <- (RM2-median(RM2, na.rm = TRUE))/sd(RM2, na.rm = TRUE) #calculating z-scores
  RM <- data.frame(GOM.median$Lab, RM1, RM2, RM1.s, RM2.s) # creating a data frame for the measurand
  # RM <- na.omit(RM) # removing all "NA"
  p <- ggplot(RM, aes(RM1, RM2, label = GOM.median.Lab))  + xlim(-5, 5) + ylim(-5,5) + 
    geom_point(aes(colour=factor(GOM.median.Lab)), size = 4)
  p <- p + xlab(y) + ylab(z) + labs(title = a) + labs(colour = "GOM.median.Lab") + mytheme + 
    geom_abline(intercept = 0, slope = 1) +  geom_abline(intercept = 2.8284, slope = 1) +  
    geom_abline(intercept = - 2.8284, slope = 1) +  geom_abline(intercept = 2.8284, slope = - 1) +  
    geom_abline(intercept = -2.8284, slope = - 1) + geom_text(data = NULL, x = -0.5, y = 2.85, label = "z = 2") + 
    geom_text(aes(colour=factor(GOM.median.Lab)), hjust=1, vjust=0) + theme(legend.position="none") #+ stat_density2d(aes(fill = ..level..), geom="polygon")
  p  + geom_errorbar(aes(ymin=RM2-RM2.s, ymax=RM2+RM2.s)) + geom_errorbarh(aes(xmin=RM1-RM1.s, xmax=RM1+RM1.s))
  
}

```

#### initial calculations with complete data set

```{r creating means}
## means over packets within lab 
meanGOM <- function(x) mean(x, na.rm=TRUE) # defining a function for further calcuations
sdGOM <- function(x) sd(x, na.rm=TRUE) # defining a function for further calcuations, here for calculating standard deviations neede for Youden plot
meanGOM.packet <- ddply(GOM, c("Lab", "Packet"), numcolwise(meanGOM)) # calculated the mean for each Packet within each Lab by calculating the mean of days and preparations lumped together
## mean over mean of packets within lab
GOM.mean <- ddply(meanGOM.packet, c("Lab"), numcolwise(meanGOM))
```
```{r creating median}
## median over packets within lab 
medianGOM <- function(x) median(x, na.rm=TRUE) # defining a function for further calcuations
medianGOM.packet <- ddply(GOM, c("Lab", "Packet"), numcolwise(medianGOM))
GOM.sd <- ddply(medianGOM.packet, c("Lab"), numcolwise(sdGOM))
## median over median of packets within lab
GOM.median <- ddply(medianGOM.packet, c("Lab"), numcolwise(medianGOM))# calculated the median for each Packet within each Lab by calculating the median of days and preparations lumped together
GOM.median <- merge(GOM.median, OKUM.methods, by="Lab")
```
```{r summary as is for indiviual measurand}
summary(GOM[[measurand]], na.rm=TRUE, digits=4) # with values without outlier removal
mean.before <- mean(GOM.mean[[measurand]], na.rm=TRUE)
median.before <- median(GOM.median[[measurand]], na.rm=TRUE)
```
The `r measurand` mean of the Lab means is **`r round(mean.before, 4)`** g/100 g  
The `r measurand` median of the Lab+Package medians is **`r round(median.before, 4)`** g/100 g  


```{r plot_lab function plot measurand vs. lab, include=FALSE}
###### function plot_lab #### 
# defined here. Automated plot of lab boxplots sorted by increasing lab mean and Horwitz function. Enter with plot_lab=('xx')
plot_lab <- function(x, type, horw = FALSE, u = FALSE) {
  element <- x
  '%p%' <- function(x, y) {as.character(paste (x, y, sep =""))}
  prep <-  'Prep.'
  method <- 'Method.'
  anal.prep <- prep %p% measurand.name
  anal.method <- method %p% measurand.name
  anal <- GOM[[element]]
  anal.prep <- GOM[[anal.prep]]
  anal.method <- GOM[[anal.method]]
  analyte <- data.frame(GOM$Lab, anal, anal.prep, anal.method )
  analyte <- na.omit(analyte)
  lab <- "U"
  if(horw=="TRUE") {
    reference.line <- median.before
         switch(type, 
         M = hor <- 1*(0.01*reference.line)^0.8495, 
         T = hor <- 10000*1*(0.01*reference.line/10000)^0.8495)
         u.lim <- reference.line + hor
         u.2lim <- reference.line + 2*hor
         l.lim <- reference.line - hor
         l.2lim <- reference.line - 2*hor
         } else  if (u == "TRUE") {
           reference.line <- median.after
             u.lim <- median.after + u2.a 
             u.2lim <- median.after + u2.a*t.value 
             l.lim <- median.after - u2.a 
             l.2lim <- median.after - u2.a*t.value 
           } else
             {u.lim <- median.after
             u.2lim <- median.after 
             l.lim <- median.after 
             l.2lim <- median.after               
             }
  p <- ggplot(reorder_by(GOM.Lab, ~anal, analyte, mean), aes(GOM.Lab, anal)) + geom_abline(intercept = reference.line, slope = 0) + geom_abline(intercept = u.2lim, slope = 0, linetype ="dotted") + geom_abline(intercept = l.2lim, slope = 0, linetype = "dotted") + geom_abline(intercept = u.lim, slope = 0, linetype ="dashed") + geom_abline(intercept = l.lim, slope = 0, linetype = "dashed")#  + geom_text( x = -Inf , y = u.2lim, label = lab) # works only for horwitz also annotations does not work
  p <- p + geom_boxplot(aes(fill=anal.method)) + geom_point(size=4) + geom_point(aes(colour=anal.prep), size = 3.5)
  p + xlab("lab") + ylab(unit) + labs(title = measurand.name) + labs(colour = "Prep")  + geom_smooth(aes(group=1)) + mytheme 
}
```

```{r plot methods, fig.width=8, fig.height=4}
plot_method(measurand) 
```
```{r plot lab plot as is, fig.width=8, fig.height=4}
plot_lab(measurand, MorT, horw=TRUE, u =FALSE) 
```

```{r fig.width=6, fig.height=6}
plot_youd(measurand.name, 'GAS', 'OKUM')
```
```{r fig.width=6, fig.height=6}
plot_youd(measurand.name, 'MUH', 'OKUM')
```
#### Mandel k barplot displays the within lab performance relative to all participating labs using the median over packages
```{r mandel k, fig.width=8, fig.height=6}
k <- with(medianGOM.packet, mandel.kh(medianGOM.packet[[measurand]], g=Lab, type="k"))     
barplot(k, las=2, col=1:4)
```

### Outlier rejection
#### Removing outliers lab based on Youden plot and Mandel's k (lab performance)

```{r outlier removal, fig.width=8, fig.height=6}
outlier <- OKUM.outlier[[measurand.name]]
#outlier <- ifelse(OKUM.outlier[[measurand.name]][1]=="NA", "0", OKUM.outlier[[measurand.name]]) # does not work for FeO
outlier <- na.omit(outlier)
# outlier<- c(16) ## defining the outlying lab here with lab# directly. Enter 0 when no outlier
leng <- length(outlier) ## counting the number of outliers for loop
for(i in seq(leng)) ##  looping
{
  GOM[[measurand]] <- ifelse(GOM$Lab==outlier[i], NA, GOM[[measurand]]) ## replacing values of outlying lab with "NA" and defining new GOM
  message("Lab ", outlier[i], " was removed")
  print(summary(GOM[[measurand]], na.rm=TRUE, digits=4))
  }

# plot_lab(measurand, MorT, horw =TRUE, u = FALSE) # replotting without outlier, M for majors and T for traces
```
## Calculating mean and median of property value

```{r calculating mean and median of property value}
summary(GOM[[measurand]], na.rm=TRUE, digits=4)
mean <- mean(tapply(GOM[[measurand]], GOM$Lab, mean, na.rm=TRUE), na.rm=TRUE) # mean of lab means after outlier removal
median.before <- median(GOM.median[[measurand]], na.rm=TRUE) # median of measurand w/o outlier removal
medianGOM.packet.after <- ddply(GOM, c("Lab", "Packet"), numcolwise(medianGOM)) # median Lab and Packets after outlier removal
## median over median of packets within lab
GOM.median.after <- ddply(medianGOM.packet.after, c("Lab"), numcolwise(medianGOM)) # creating a new table of median of labs after outlier removal for measurand but for entire table (all measurands)
GOM.median.after <- merge(GOM.median.after, OKUM.methods, by="Lab")
median.after <- median(GOM.median.after[[measurand]], na.rm=TRUE) # median of the individual measurand after outlier removal
## calculating method parameters
#  '%p%' <- function(x, y) {as.character(paste (x, y, sep =""))}
prep <-  'Prep.'
method <- 'Method.'
anal.prep <- prep %p% measurand.name
anal.method <- method %p% measurand.name
anal <- GOM.median.after[[measurand]]
anal.prep <- GOM.median.after[[anal.prep]]
anal.method <- GOM.median.after[[anal.method]]
analyte <- data.frame(GOM.median.after$Lab, GOM.median.after$names, anal, anal.prep, anal.method )
analyte <- na.omit(analyte)

  bymethod.n <- ddply(analyte, c("anal.method"), summarise, 
                      N=length(anal), 
                      mean = round(mean(anal), 3), 
                      median = round(median(anal), 3), 
                      sd = round(sd(anal),3), 
                      se = round(sd/sqrt(N),3))

```

## Measurement uncertainty estimations of property value
### before outlier removal

```{r measurement uncertainty of data before outlier rejection, fig.width=6, fig.height=6}
medianGOM.packet$Lab <- as.factor(medianGOM.packet$Lab) # using only the median of the 3 packages per lab
medianGOM.packet$Packet <- as.factor(medianGOM.packet$Packet)
anal <- medianGOM.packet[[measurand]]
DF.lme <- data.frame(medianGOM.packet$Lab, medianGOM.packet$Packet, medianGOM.packet[[measurand]])
DF.lme <- na.omit(DF.lme)
names(DF.lme) <- c("Lab", "Packet", "measurand")
GOM.lme <- lme(measurand ~ 1, random = ~ 1|Lab/Packet, data=DF.lme) # linear model with random effects
sL2 <- varcomp(GOM.lme, FALSE, FALSE)[[1]] # between-laboratory variance
sbb2 <- varcomp(GOM.lme, FALSE, FALSE)[[2]] # between bottle variance
sr2 <- varcomp(GOM.lme, FALSE, FALSE)[[3]] # repeatability standard deviation
n.p <- dim(DF.lme)[1] # number of obervations
p <- length(unique(DF.lme$Lab)) # haven't found a better way how to extract the number of labs (number of groups)
r <- length(unique(DF.lme$Packet))
u1 <- sqrt(sL2/p+sbb2/p/r+sr2/p/r/4) # calculating the standard uncertainty of characterization
u2 <- attr(GOM.lme$fixDF,"varFixFact") # gives the same results as u, amazing!
plot(DF.lme)
```
### after outlier removal

```{r measurement uncertainty of data after outlier rejection, fig.width=6, fig.height=6}
medianGOM.packet.after$Lab <- as.factor(medianGOM.packet$Lab) # using only the median of the 3 packages per lab
medianGOM.packet.after$Packet <- as.factor(medianGOM.packet$Packet)
anal <- medianGOM.packet.after[[measurand]]
DF.lme <- data.frame(medianGOM.packet.after$Lab, medianGOM.packet.after$Packet, medianGOM.packet.after[[measurand]])
DF.lme <- na.omit(DF.lme)
names(DF.lme) <- c("Lab", "Packet", "measurand")
GOM.lme <- lme(measurand ~ 1, random = ~ 1|Lab/Packet, data=DF.lme) # linear model with random effects
sL2.a <- varcomp(GOM.lme, FALSE, FALSE)[[1]] # between-laboratory variance
sbb2.a <- varcomp(GOM.lme, FALSE, FALSE)[[2]] # between bottle standard deviation
sr2.a <- varcomp(GOM.lme, FALSE, FALSE)[[3]] # repeatability standard deviation
n.p <- dim(DF.lme)[1] # number of obervations
p <- length(unique(DF.lme$Lab)) # haven't found a better way how to extract the number of labs (number of groups)
r <- length(unique(DF.lme$Packet))
t.value <- qt(0.975,df=p-1)
u1.a <- sqrt(sL2.a/p+sbb2.a/p/r+sr2.a/p/r/4) # calculating the standard uncertainty of characterization
u2.a <- attr(GOM.lme$fixDF,"varFixFact") # gives the same results as u1, amazing!
plot(DF.lme)
```
 
 
 
### tests for normal distribution
```{r QQ plot, fig.width=4.5, fig.height=5}
qqnorm(GOM.median.after[[measurand]])
qqline(GOM.median.after[[measurand]])
```
```{r without PP}
analyte.noPP <- subset(analyte, analyte$anal.prep!="PP") # Removing all PP preparations for comparison reasons
median.after.noPP <- median(analyte.noPP$anal)
```

## Comparisons of property value calculations for `r refmat` and measurand `r measurand.name` 
### before and after removal of outlying lab(s) # `r outlier`
type |  before | after | unit 
:----| :------- | :-------- |:------ | :--------
**mean** |`r round(mean.before, 4)`  |`r round(mean, 4)` |`r unit`
**median** | `r round(median.before, 4)` | **`r round(median.after, 4)`**| `r unit`
**median w/o PP** | - | **`r round(median.after.noPP, 4)`**| `r unit`


### performance characteristics for `r refmat` and measurand `r measurand.name` 
type | variance before | variance after outlier rejection 
:----| :------- | :-----------
between-laboratory variance | `r sL2` | `r sL2.a` 
between-bottle variance | `r sbb2` | `r sbb2.a`
repeatability variance | `r sr2` | `r sr2.a` 

### mesasurment uncertainty
type |  before |  after outlier rejection | alternative calculation
:----| :------- | :----------- | :--------
standard uncertainty | `r u2` | `r u1.a` | `r u2.a`
expanded standard uncertainty |-|-| **`r u2.a*t.value`** *k* = `r round(t.value, 2)`

```{r exporting the data}
outlier <- ifelse(outlier=="0", "X", outlier)
df <- data.frame(Sys.Date(), refmat, measurand.name, mean.before, mean, median.before, median.after, median.after.noPP, unit, sL2.a, sbb2.a, sr2.a, u1.a, u2.a, u2.a*t.value, t.value, outlier, p)
# names(df) <- c("date", "RM", "measurand", "mean.before", "mean", "median.before", "median.after", "median.after.noPP", "unit", "sL2.a", "sbb2.a", "sr2.a", "u", "u.alternative", "U", "t.value", "outlier", "labs remaining") # needed only the first time
# write.table(df, "df.txt", row.names=FALSE) # needed only the first time
write.table(df, "df.txt", row.names=FALSE, append=TRUE, col.names=FALSE)
```
```{r plotting the method medians, fig.width=8, fig.height=4}
reference.line <- median.after
u.Ulim <- median.after + u2.a*t.value
l.Ulim <- median.after - u2.a*t.value
ggplot(bymethod.n, aes(x=anal.method, y=median))+geom_point(size=4)+geom_errorbar(aes(ymin=median-se, ymax=median+se), width=0.05)+ geom_abline(intercept = reference.line, slope = 0) + geom_abline(intercept = u.Ulim, slope = 0, linetype ="dotted") + geom_abline(intercept = l.Ulim, slope = 0, linetype = "dotted") + mytheme
```
### Final plot without outliers with u and U

```{r fig.width=8, fig.height=4}
plot_lab(measurand, MorT, horw = FALSE, u = TRUE)
```
## Comparisons of method parameters based on median of the Lab+packet medians after outlier removal
```{r method comparison output}
print(bymethod.n)
```
```{r for output, include=FALSE }
outlier.type.name <- measurand.name %p% ".outlier.type" # defining if outlier is selected ("Y" or NA)
outlier.dist.type <- measurand.name %p% ".dist.type" # defining if outlier is based on Y = Youden plot, M = Madel's k, P = pressed powder pellet, D = detection limit 
out.measurand <- data.frame(OKUM.outlier[[measurand.name]], OKUM.outlier[[outlier.type.name]], OKUM.outlier[[outlier.dist.type]])
# out.measurand <- cbind(OKUM.outlier[[measurand.name]], OKUM.outlier[[outlier.type.name]])
names(out.measurand) <- c("outlier.lab","outlier.type", "outlier.dist" )
property.value.dist.type <- out.measurand[1,3]
property.value <- ifelse(property.value.dist.type == "median", median.after, mean)
outY <- subset(out.measurand, outlier.type =="Y", select= c(outlier.lab))
outY <- outY$outlier.lab
outM <- subset(out.measurand, outlier.type =="M", select= c(outlier.lab))
outM <- outM$outlier.lab
outP <- subset(out.measurand, outlier.type =="P", select= c(outlier.lab))
outP <- outP$outlier.lab
outD <- subset(out.measurand, outlier.type =="D", select= c(outlier.lab))
outD <- outD$outlier.lab
```


### final result based on median for property value and measurement uncertainty based on variance components
The **`r measurand.name`** `r property.value.dist.type` of the Lab+packet `r property.value.dist.type` after outlier removal is **`r round(property.value, 4)`** `r unit`   
The expanded standard uncertainty for the assigned value of `r measurand.name` is **`r u2.a*t.value`** with k = `r round(t.value, 2)`   
excluded labs for `r measurand.name` is/are `r outlier`  
labs remaining for calculations `r p`  
#### comments


outliers:|lab#
--------|-----
based on Youden plots:|`r outY`
based on Mandel's k performance:|`r outM`
based on PP-XRF:|`r outP`
based on detection limit:| `r outD`
